Supplementary Information for RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2014-03-31 doc/bioinformatics2014/document.tex
Update the manuscript

We use a two-sided posterior difference test using control and case samples to test for somatic mutations. We consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g. loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g. homozygous somatic mutation). The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$. We typically set the hreshold $ \tau $ to zero.

@@

We use a one-sided posterior distribution test using a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,

@@

\subsection{HCC1187 Sequence Data}
\subsubsection*{Experimental process.}
The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. Matched normal cells were derived from lymphoblastoid cells from peripheral blood. 
\subsubsection*{Computational process.}
Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}. The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
%Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx

@@

We performed the posterior distribution test for somatic mutations on the synthetic data set to identify mutations in the haploid DNA sequence. We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.

@@

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
%\afterpage{\clearpage}
\subsubsection*{Sensitivity/Specificity Comparison}

@@

\begin{figure*}[!hbpt]
\begin{center}
\includegraphics[width=0.85\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\label{fig:brca_MAF}
\end{center}
\end{figure*}

@@

RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. There were eight germline variants and eight somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations. Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation. Positions chr7:154754371 and chr7:154758813 were considered loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

>>>>

We use a two-sided posterior difference test with control and case paired samples to identify somatic mutations. We consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g. loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g. homozygous somatic mutation). The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$. We typically set the hreshold $ \tau $ to zero.

@@

We use a one-sided posterior distribution test with a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,

@@

\subsection{HCC1187 Sequence Data}
\subsubsection*{Experimental process.}
The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. Matched normal cells were derived from lymphoblastoid cells from peripheral blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
\subsubsection*{Computational process.}
 The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
%Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx

@@

We performed the posterior difference test to identify mutations in the haploid synthetic data. We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.

@@

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
% from folder 'rvd2\results\2013-09-19_operating_characteristics'

\subsubsection*{Sensitivity/Specificity Comparison}

@@

\begin{figure*}[!hbpt]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\label{fig:brca_MAF}
\end{center}
\end{figure*}

@@

RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. There were eight germline variants and eight somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations. Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation. Positions chr7:154754371 and chr7:154758813 appears to be loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

### 2014-03-31 doc/bioinformatics2014/supplementary_info.tex
Update the manuscript

\begin{figure}[!bpth]
    \centering
    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
    \quad
    \quad
    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
    \caption{(A) RVD2 Graphical Model. (B) Example to illustrate the proposal graphical model and generative process. Here we want to generate a simulation sample with 3 replicates, and each replicate is a 5bp long sequence. Therefore, we have $ J = 5 $, $N = 3$. The hyperparameters in orange, $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$, $n \in \RR^{5 \times 3}$ are set. (1) Sample the local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$. (2) Use $\mu $ and $ M $ to sample $\theta \in \RR^{5 \times 3}$ by drawing random samples $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$. (3) Use $\theta \in \RR^{5 \times 3}$ and $n \in \RR^{5 \times 3}$ to sample $r \in \RR^{5 \times 3}$ by drawing random numbers $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.} 
    \label{fig:brca_MAF}%
\end{figure}

@@

RVD2 achieved MCC value 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260. This indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
%%%%%%%%%%%%%%%%%
% Appendix D: Estimated minor allele fraction for called variants in synthetic gene.
%%%%%%%%%%%%%%%%%
\section{Estimated MAF for called variants in synthetic gene.}
Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}
% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

@@

Samtools mpileup and GATK can accept multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup. VarScan2-mpileup takes multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm. Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.

\section{histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.} 
\label{fig:hist_pos}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
\label{fig:hist_neg}
\end{center}
\end{figure}

>>>>

\begin{figure}[!bpth]
    \centering
    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
    \quad
    \quad
    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
    \caption{(A) RVD2 Graphical Model. (B) Example to illustrate the proposal graphical model and generative process. Here we want to generate a simulation sample with 3 replicates, and each replicate is a 5bp long sequence. Therefore, we have $ J = 5 $, $N = 3$. The hyperparameters in orange, $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$ are set. $n \in \RR^{5 \times 3}$ in the generative process is given as well. The generative process is: (1) Sample the local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$. (2) Use $\mu $ and $ M $ to sample $\theta \in \RR^{5 \times 3}$ by drawing random samples $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$. (3) Use $\theta \in \RR^{5 \times 3}$ and $n \in \RR^{5 \times 3}$ to sample $r \in \RR^{5 \times 3}$ by drawing random numbers $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.} 
    \label{fig:brca_MAF}%
\end{figure}

@@

RVD2 achieved MCC value 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260. This indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
%%%%%%%%%%%%%%%%%
% Appendix D: Estimated minor allele fraction for called variants in synthetic gene.
%%%%%%%%%%%%%%%%%
\section{Estimated MAF for called variants in synthetic gene.}
Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}
% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

@@

Samtools mpileup and GATK can accept multiple tumor replicates for variant calling, so we fed six bam files from each case replicate group to mpileup. VarScan2-mpileup takes multiple tumor-normal pair replicates so we passed six pair replicates to each algorithm. Varscan2-somatic, strelka and muTect do not accept replicate data for the normal or tumor bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.

\section{histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.} 
\label{fig:hist_pos}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_neg.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
\label{fig:hist_neg}
\end{center}
\end{figure}
